Supplemental material

### 2017-11-12 output/breed_paper.Rnw
Updates with full random slopes on yrs_in_trial.
Updated supplemental analysis.

\title{Rice yield improvements through plant breeding are offset by
  inherent yield declines over time}
\author[ucd]{Matthew B. Espe\corref{corresauthor}}
\cortext[corresauthor]{Corresponding author} \ead{mespe@ucdavis.edu}
\author[ucd]{David Mackill}
\author[ucd]{Jim E. Hill}
\author[ucce2]{Michelle Leinfelder-Miles}
\author[ucce]{Luis A. Espino}
\author[ucce]{Randall Mutters}
\author[ucd]{Chris van Kessel}
\author[ucd]{Bruce A. Linquist}

@@

There was general agreement between the estimates from the Bayesian
and classical approaches.  The Bayesian estimates are presented here
due to the inherent flexibility of Bayesian inference, though the
classical estimates can be found in the supplemental
material. Comparing the cultivars' yield during each's initial time
(average 3 years) in the experimental trials just before being
released showed that across cultivars, yields ranged from
\Sexpr{round(plot_data["M207", "mean"], 2)} (M-207) to
\Sexpr{round(plot_data["M205", "mean"], 2)} Mg ha\tsups{-1} (M-205)
(Fig \ref{fig:RES_year}. Evaluating these initial yields over time
indicates there was no evidence of genetic yield improvements during
the 30-year study period (Fig. \ref{fig:RES_year}). Individual exceptions to this were
two cultivars (M-204 and M-205) that had higher yields than the
others.

@@

Looking at all the cultivars that have been released, there is no
conclusive evidence of increased genetic yield potential in medium
grain rice cultivars released by CCRRF's breeding improvement program
over the period \Sexpr{min(mod_data[,'year'])} to
\Sexpr{max(mod_data[,'year'])}, which has stayed relatively constant
at \Sexpr{round(mean(mod_data[,"yield_kg"])/1000, 1)} Mg
ha\tsups{-1}. Since the estimated yield improvement spans from
\Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]} kg ha\tsups{-1} year\tsups{-1}
(Fig. \ref{fig:RES_year}; Table \ref{tbl:coef}),
both increases and decreases in the genetic yield potential over the
study period are plausible. The estimated yield
improvement of \Sexpr{only_release_coef} kg ha\tsups{-1}
year\tsups{-1} (Fig. \ref{fig:wo_erosion}) that does not account for
yield erosion is lower than the 37 to 60 kg ha\tsups{-1}
year\tsups{-1} estimated by~\citet{Omar2014} in side-by-side
comparisons. Rather, this analysis suggests that the CCRRF breeding
program has maintained but not substantially improved the yield
potential in the face of changing biotic and abiotic pressures. This
conclusion holds true even when evaluating the cultivars which have
been breed for broad adaption and high yield potential (M-202, M-206
and M-105). Yields of these cultivars at time of release have all
yielded between
\Sexpr{tmp=mod_data[mod_data[,"yrs_in_trial"]<=0&mod_data[,"id"]%in%c("M202","M206","M105"),];round(min(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
and
\Sexpr{round(min(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
Mg ha\tsups{-1} with no indication of yields increasing
(Fig. \ref{fig:first_yrs}). While these yields have not increased,
M-205 did have higher yield potential; however, this cultivar is not
broadly adapted due to poor cold tolerance.

>>>>

\title{Rice yield improvements through plant breeding are offset by
  inherent yield declines over time}
\author[ucd]{Matthew B. Espe\corref{corresauthor}}
\cortext[corresauthor]{Corresponding author} \ead{mespe@ucdavis.edu}
\author[ucd]{Jim E. Hill}
\author[ucce2]{Michelle Leinfelder-Miles}
\author[ucce]{Luis A. Espino}
\author[ucce]{Randall Mutters}
\author[ucd]{David Mackill}
\author[ucd]{Chris van Kessel}
\author[ucd]{Bruce A. Linquist}

@@

There was general agreement between the estimates from the Bayesian
and classical approaches.  The Bayesian estimates are presented here
due to the inherent flexibility of Bayesian inference, though the
classical estimates can be found in the supplemental
material. Comparing the cultivars' yield during each's initial time
(average 3 years) in the experimental trials just before being
released showed that across cultivars, yields ranged from
\Sexpr{round(plot_data["M207", "mean"], 2)} (M-207) to
\Sexpr{round(plot_data["M205", "mean"], 2)} Mg ha\tsups{-1} (M-205)
(Fig \ref{fig:first_yrs}. Evaluating these initial yields over time
indicates there was no evidence of genetic yield improvements during
the 30+ year study period (Fig. \ref{fig:RES_year}). Individual exceptions to this were
two cultivars (M-204 and M-205) that had higher yields than the
others.

@@

Looking at all the cultivars that have been released, there is no
conclusive evidence of increased genetic yield potential in medium
grain rice cultivars released by CCRRF's breeding improvement program
over the period \Sexpr{min(mod_data[,'year'])} to
\Sexpr{max(mod_data[,'year'])}, which has stayed relatively constant
at \Sexpr{round(mean(mod_data[,"yield_kg"])/1000, 1)} Mg
ha\tsups{-1}. Since the estimated yield improvement spans from
\Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]} kg ha\tsups{-1} year\tsups{-1}
(Fig. \ref{fig:RES_year}; Table \ref{tbl:coef}),
both increases and decreases in the genetic yield potential over the
study period are plausible. The estimated yield
improvement of \Sexpr{only_release_coef} kg ha\tsups{-1}
year\tsups{-1} (Fig. \ref{fig:wo_erosion}) that does not account for
yield erosion is lower than the 37 to 60 kg ha\tsups{-1}
year\tsups{-1} estimated by~\citet{Omar2014} in side-by-side
comparisons. Rather, this analysis suggests that the CCRRF breeding
program has maintained but not substantially improved the yield
potential in the face of changing biotic and abiotic pressures. This
conclusion holds true even when evaluating the cultivars which have
been breed for broad adaption and high yield potential (M-202, M-206
and M-105). Yields of these cultivars at time of release have all
yielded between
\Sexpr{tmp=mod_data[mod_data[,"yrs_in_trial"]<=0&mod_data[,"id"]%in%c("M202","M206","M105"),];round(min(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
and
\Sexpr{round(max(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
Mg ha\tsups{-1} with no indication of yields increasing
(Fig. \ref{fig:first_yrs}). While these yields have not increased,
M-205 did have higher yield potential; however, this cultivar is not
broadly adapted due to poor cold tolerance.

### 2017-11-12 output/freq.Rmd
Commit before inverting the sections in results.
Updated supplemental analysis.

---
output: pdf_document
title: "Supplemental material"
date: "Oct. 1, 2017"
geometry: margin=1in
---

```{r setup, echo = FALSE}
load("../data/model_data.rda")
library(lme4)

@@

This document replicates the analysis in "Rice yield improvements
through plant breeding are offset by inherent yield declines over
time" (Espe et al, 2017) using classical/frequentist statistics to
verify the results are consistent regardless of statistical method.

The complete data, code, and manuscript is available from \url{https://github.com/mespe/rice_advances}.

# Model excluding yield erosion

```{r}
no_erosion = lmer(yield_kg ~ release_yr +
                      (1|id) + (1|year), data = mod_data)

summary(no_erosion)
confint(no_erosion, parm = c("release_yr"),
        method = "boot", level = 0.9)
```

# Model including yield erosion

```{r}
erosion = lmer(yield_kg ~ release_yr + yrs_in_trial +
                   (1|id) + (1|year), data = mod_data)

summary(erosion)
confint(erosion, parm = c("release_yr","yrs_in_trial"),
        method = "boot", level = 0.9)
```

>>>>

---
output: pdf_document
title: "Supplemental material"
date: \today{}
geometry: margin=1in
header-includes:
   - \usepackage{dcolumn}
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
load("../data/model_data.rda")
library(lme4)
library(stargazer)
library(xtable)
```

This document replicates the analysis in "Rice yield improvements
through plant breeding are offset by inherent yield declines over
time" (Espe et al, 2017) using classical/frequentist mixed-effects
models. The purpose of this document to verify the results are
consistent regardless of statistical method used (i.e., a "sanity
check"). As the table below show, the estimates are indeed similar. As
such, please refer to the main manuscript for discussion of the model
estimates.

For each model, the structure is the same as used in the Bayesian
analysis (see main manuscript for details). The models were fit in R
using the 'lme4' package (Bates et al., 2015). The complete data,
code, and manuscript is available from
[https://github.com/mespe/rice_advances](https://github.com/mespe/rice_advances).

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "asis"}
no_erosion = lmer(yield_kg ~ release_yr +
                      (1|id) + (1|year), data = mod_data)

stargazer(no_erosion, title = "Model excluding yield erosion.",
          ci = TRUE, ci.level = 0.90, float = TRUE,
          covariate.labels = c("Year of release", "Intercept"),
          dep.var.caption = "Response",
          dep.var.labels = "Yield ( kg/ha )",
          digits = 1, header = FALSE
          )

@@

```{r, echo = FALSE, results = "asis"}
erosion = lmer(yield_kg ~ release_yr + yrs_in_trial +
                   (1|id) + (1|year), data = mod_data)

stargazer(erosion, title = "Model including yield erosion.",
          ci = TRUE, ci.level = 0.90, float = TRUE,
          covariate.labels = c("Year of release", "Yield erosion","Intercept"),
          dep.var.caption = "Response",
          dep.var.labels = "Yield ( kg/ha )",
          digits = 1, header = FALSE
          )
```

```{r, message = FALSE, warning = FALSE, results = "asis", echo = FALSE}
tt = anova(erosion, no_erosion)
row.names(tt) = c("Excluding yield erosion", "Including yield erosion")
print(xtable(tt,
             caption = "Model comparison"), comment = FALSE,
      caption.placement = "top")
